Windowed residual-minimization model reduction for dynamical systems

### 2019-10-03 section4/spatial_numerical.tex

\subsection{\spatialAcronym: Direct Solution Approach}\label{sec:direct} 
Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.

@@

\subsubsection{Linear Multistep Schemes}

@@

\begin{remark}
For the limiting case where the window size is equivalent to the time-step size, $\DeltaSlabArg{n} \equiv \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$, and uniform 
quadrature weights are used, \methodAcronym\ recovers LSPG.
\end{remark} 
%%%%%%%%%%%%%%%%COLLOCATION%%%%%%%
\begin{comment}
\subsubsection{Collocation (Runge--Kutta) Methods}
Collocation methods comprise another popular technique for temporal discretization. 
Collocation methods represent the FOM state over each time-step instance $[\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}]$ as,
\begin{equation}
\stateFOMArgnt{n}(t) \approx  \collocMatArg{n}{i} \collocBasisArg{n}{i}(t), \qquad t \in [\timeWindowArg{n}{i-1}, \timeWindowArg{n}{i} ],
\end{equation}
with the
continuity and boundary constraints, 
\begin{equation}\label{eq:colloc_constraint}
\collocMatArg{n}{i} \collocBasisArg{n}{i}(\timeWindowArg{n}{i-1}) = 
\begin{cases}
\collocMatArg{n}{i-1}\collocBasisArg{n}{i-1}(\timeWindowArg{n}{i-1}) & n = 1,\ldots,\nslabs, \; i=2,\ldots,\nstepsArg{n},\\ 
\collocMatArg{n-1}{\nstepsArg{n-1}}\collocBasisArg{n-1}{\nstepsArg{n-1}}(\timeWindowArg{n-1}{\nstepsArg{n-1}}) & n = 2,\ldots,\nslabs, \; i=1,  \\
\stateFOMIC & n=1, \;  i=1.
\end{cases}
\end{equation}
%\begin{equation}\label{eq:continuity_constraint}
%\genCollocMatArg{n}{i} \collocBasisArg{n}{i}(\timeWindowArg{n}{i-1}) = 
%%\genstatecollocArg{n}{i}{\timeWindowArg{n}{i-1}} =
%%\genstatecollocArg{n}{i-1}{\timeWindowArg{n}{i-1}} , \qquad  i =
%%2,\ldots,\nstepsArg{n}. 
%\begin{cases} \genstatecollocArg{n}{i-1}{\timeWindowArg{n}{i-1}} & i =
%2,\ldots,\nstepsArg{n} \\
%\genstatecollocArg{n-1}{\nstepsArg{n}}{\timeWindowArg{n}{i-1}} & i= 1. \\
%\end{cases} 
%\end{equation} 
In the above, $\collocMatArg{n}{i}  \in \RR{\fomdim \times \collocOrderArg{n}{i}}$ are \textit{space--time coordinates}
over the $i$th time-step instance of the $n$th window, 
%$\genstatecollocArgnt{n}{i} : [\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}] \mapsto \RR{\fomdim}$ is the collocated 
%approximation to the state over the $i$th time-step instance within the $n$th slab
and $\collocBasisArg{n}{i} :
[\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}] \mapsto \RR{\collocOrderArg{n}{i}}$ comprise the
$\collocOrderArg{n}{i}$ 
\textit{polynomial basis functions} over the $i$th time-step instance of the $nth$ window. 
%While a variety of temporal basis functions are possible,
%polynomial basis functions are the most commonly employed and as such we consider these here.

@@

\begin{remark} %The presented collocation formulation corresponds to minimizing the
%weighted O$\Delta$E residual over the time slab, in where the O$\Delta$E
%results from temporal
%discretization of the FOM ODE according to the temporal basis and collocation points. For 
%the case where the temporal basis functions are
%polynomials, the resulting temporal discretization schemes belongs to the family of Runge--Kutta
%schemes~\cite{1}. 
Collocation methods belong to the family of Runge--Kutta schemes.
Alternative temporal discretization schemes, such as
linear multistep methods, can be obtained by modifying the representation of
the state and the collocation points~\cite{ajams2015362}.
\end{remark}
\end{comment}
\subsubsection{Solution to the Nonlinear Least-Squares Problem through the Gauss--Newton Method}
Discretization through linear multistep methods (as well as other techniques) 
results in a nonlinear least-squares problem.
A variety of algorithms exist for solving nonlinear  
least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method.  
The numerical experiments presented in this work consider the Gauss--Newton method, and as such we outline this approach here.

@@

\subsection{Spatial Trial Space: Indirect Solution Approach}\label{sec:indirect}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler-Lagrange equations. The
system given by the Euler Lagrange equations~\eqref{eq:lspg_continuous}-\eqref{eq:lspg_adjoint} comprises a coupled two-point boundary value
problem. Similar to the direct solution approach, a variety of techniques have
been devised to solve two-point boundary value problems of this type. These
techniques include shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward-backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.

@@

\begin{algorithm} \caption{\spatialAcronym\ trial subspace: algorithm for the FBSM over the $n$th window.} \label{alg:st_iter} 
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $tol$; damping factor, $\rho$; growth factor, $\fbsmGrowth$; decay
factor, $\fbsmDecay$} 
\Output{Stationary point, $\genstateArgnt{n}$ }
\textbf{Online Steps:}\\ 
$\genstate^n_0 \leftarrow \bz$ \Comment{Set initial guess for state} \\
$\controllerArgnt{n} \leftarrow \boldsymbol 0$ \Comment{Set initial
guess for costate}\\ 
$\genstateArgnt{n}_1 \solveAlgo \basisspace^T \stweightingMatArg{n}
\basisspace \genstateDotArgnt{n}_1  -  \basisspace^T \stweightingMatArg{n}
\velocity(\veloargsromntArg{1}) =  \controllerArgnt{n}$ 
% Eq.~\ref{eq:lspg_continuous} with $\adjoint^n(t) = \controllerArg{n}{t}$ to
% obtain $\genstate_0^{n}(t)$ 
\Comment{Solve~\eqref{eq:lspg_continuous} with initial guess to
controller}\\ 
%Set: $\genstate^n_1 = \genstate$ \Comment{Set state} \\
%$\objectiveArg{n}({\genstate_1^n(t)})$
%\Comment{Evaluate objective function}\\
$i \leftarrow 1$ \Comment{Set counter}\\
\While{$tol \le \int_{\timeStartArg{n}}^{\timeEndArg{n}} \norm{\genstate^n_{i}(t) - \genstate^n_{i-1}(t) }dt $}{
\small{
\begin{multline*}
\adjointArgnt{n} \underset{\text{solve}}{\leftarrow} 
\frac{d}{dt} \adjoint^n  + \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\basisspace \genstate^n_i + \stateIntercept,t) \bigg]^T \stweightingMatArg{n} \basisspace \mass^{-1} \adjoint^n = \\ -\bigg[\basisspace^T \bigg[ \frac{\partial \velocity}{\partial \stateyDiscrete} ( \basisspace \genstate^n_i + \stateIntercept,t) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} -   \basisspace \mass^{-1} \basisspace^T \bigg)  \stweightingMatArg{n} \bigg( \basisspace \dot{\genstate}_i^n   -   \velocity( \basisspace \genstate_i^n  +\stateIntercept,t) \bigg) \bigg] 
\end{multline*} }
\Comment{Solve~\eqref{eq:lspg_adjoint} to obtain guess to costate state} \\
%Solve Eq.~\ref{eq:lspg_adjoint} with $\genstate^n(t) = \genstate^n_i(t)$ to obtain $\adjoint^n(t)$ \\
$\controllerArgnt{n}  \leftarrow \rho \controllerArgnt{n} + (1 - \rho) \adjointArgnt{n}$ \Comment{Weighted update to costate}\\
$i \leftarrow i+1$ \Comment{Update Counter}\\
$\genstateArgnt{n}_i \solveAlgo \basisspace^T \stweightingMatArg{n} \basisspace \genstateDotArgnt{n}_i   -  \basisspace^T \stweightingMatArg{n} \velocity(\basisspace \genstate^n_i + \stateIntercept,t) =  \controllerArgnt{n} $
\Comment{Solve~\ref{eq:lspg_continuous} with updated guess to costate}\\
%$\objectiveArg{n}({\genstate_i^n(t)})$ \Comment{Evaluate objective function}\\
\uIf{ $\objectiveArg{n}({\genstate_i^n(t)}) \le \objectiveArg{n}({\genstate_{i-1}^n(t)})$}
{
$\rho \leftarrow \text{min}(\rho \fbsmGrowth,1)$ \Comment{Grow the damping factor}\\
}
\Else{
$\rho \leftarrow \frac{\rho }{ \fbsmDecay}$ \Comment{Shrink the damping factor}\\ 
$\genstate_i^{n} \leftarrow  \genstate_{i-1}^{n}$ \Comment{Reset state to value at previous iteration}
}
}
Return converged solution, $\genstateArgnt{n}= \genstateArgnt{n}_i$
\end{algorithm}
\subsubsection{Numerical Considerations for the Forward and Backwards Problems}
The FBSM requires solving~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}, both of which are defined at the time-continuous level. 
The numerical implementation of the FBSM requires two main ingredients: (1.) temporal discretization schemes for the forward and backward problems and (2.) 
an efficient method for computing action of the Jacobian transpose on a vector.

>>>>

\subsection{\spatialAcronym: direct solution approach}\label{sec:direct} 
Direct approaches solve~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective function in time.
The minimization problem is then reformulated as a nonlinear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, orthogonal polynomial (spectral  method) 
approaches, and genetic algorithms.

@@

\subsubsection{Linear multistep schemes}

@@

\begin{remark}
For the limiting case where the window size is equivalent to the time-step size, $\DeltaSlabArg{n} \equiv \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$, and uniform 
quadrature weights are used, \methodAcronym\ recovers LSPG.
\end{remark} 
%%%%%%%%%%%%%%%%COLLOCATION%%%%%%%
\begin{comment}
\subsubsection{Collocation (Runge--Kutta) methods}
Collocation methods comprise another popular technique for temporal discretization. 
Collocation methods represent the FOM state over each time-step instance $[\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}]$ as,
\begin{equation}
\stateFOMArgnt{n}(t) \approx  \collocMatArg{n}{i} \collocBasisArg{n}{i}(t), \qquad t \in [\timeWindowArg{n}{i-1}, \timeWindowArg{n}{i} ],
\end{equation}
with the
continuity and boundary constraints, 
\begin{equation}\label{eq:colloc_constraint}
\collocMatArg{n}{i} \collocBasisArg{n}{i}(\timeWindowArg{n}{i-1}) = 
\begin{cases}
\collocMatArg{n}{i-1}\collocBasisArg{n}{i-1}(\timeWindowArg{n}{i-1}) & n = 1,\ldots,\nslabs, \; i=2,\ldots,\nstepsArg{n},\\ 
\collocMatArg{n-1}{\nstepsArg{n-1}}\collocBasisArg{n-1}{\nstepsArg{n-1}}(\timeWindowArg{n-1}{\nstepsArg{n-1}}) & n = 2,\ldots,\nslabs, \; i=1,  \\
\stateFOMIC & n=1, \;  i=1.
\end{cases}
\end{equation}
%\begin{equation}\label{eq:continuity_constraint}
%\genCollocMatArg{n}{i} \collocBasisArg{n}{i}(\timeWindowArg{n}{i-1}) = 
%%\genstatecollocArg{n}{i}{\timeWindowArg{n}{i-1}} =
%%\genstatecollocArg{n}{i-1}{\timeWindowArg{n}{i-1}} , \qquad  i =
%%2,\ldots,\nstepsArg{n}. 
%\begin{cases} \genstatecollocArg{n}{i-1}{\timeWindowArg{n}{i-1}} & i =
%2,\ldots,\nstepsArg{n} \\
%\genstatecollocArg{n-1}{\nstepsArg{n}}{\timeWindowArg{n}{i-1}} & i= 1. \\
%\end{cases} 
%\end{equation} 
In the above, $\collocMatArg{n}{i}  \in \RR{\fomdim \times \collocOrderArg{n}{i}}$ are \textit{space--time coordinates}
over the $i$th time-step instance of the $n$th window, 
%$\genstatecollocArgnt{n}{i} : [\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}] \mapsto \RR{\fomdim}$ is the collocated 
%approximation to the state over the $i$th time-step instance within the $n$th slab
and $\collocBasisArg{n}{i} :
[\timeWindowArg{n}{i-1},\timeWindowArg{n}{i}] \mapsto \RR{\collocOrderArg{n}{i}}$ comprise the
$\collocOrderArg{n}{i}$ 
\textit{polynomial basis functions} over the $i$th time-step instance of the $nth$ window. 
%While a variety of temporal basis functions are possible,
%polynomial basis functions are the most commonly employed and as such we consider these here.

@@

\begin{remark} %The presented collocation formulation corresponds to minimizing the
%weighted O$\Delta$E residual over the time slab, in where the O$\Delta$E
%results from temporal
%discretization of the FOM ODE according to the temporal basis and collocation points. For 
%the case where the temporal basis functions are
%polynomials, the resulting temporal discretization schemes belongs to the family of Runge--Kutta
%schemes~\cite{1}. 
Collocation methods belong to the family of Runge--Kutta schemes.
Alternative temporal discretization schemes, such as
linear multistep methods, can be obtained by modifying the representation of
the state and the collocation points~\cite{ajams2015362}.
\end{remark}
\end{comment}
\subsubsection{Solution to the nonlinear least-squares problem through the Gauss--Newton method}
Discretization through linear multistep methods (as well as other techniques) 
results in a nonlinear least-squares problem.
A variety of algorithms exist for solving nonlinear  
least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method.  
The numerical experiments presented in this work consider the Gauss--Newton method, and as such we outline this approach here.

@@

\subsection{Spatial trial space: indirect solution approach}\label{sec:indirect}
In contrast to the direct approach,
indirect methods ``indirectly" solve the minimization
problem~\eqref{eq:tclsrm} by solving the Euler-Lagrange equations. The
system given by the Euler Lagrange equations~\eqref{eq:lspg_continuous}-\eqref{eq:lspg_adjoint} comprises a coupled two-point boundary value
problem. Similar to the direct solution approach, a variety of techniques have
been devised to solve two-point boundary value problems of this type. These
techniques include shooting methods, multiple shooting
methods~\cite{multiple_shooting}, and the forward-backward sweep
method~\cite{fbs} (FBSM).  This work explores using the FBSM.

@@

\begin{algorithm} \caption{\spatialAcronym\ trial subspace: algorithm for the FBSM over the $n$th window.} \label{alg:st_iter} 
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $tol$; damping factor, $\rho$; growth factor, $\fbsmGrowth$; decay
factor, $\fbsmDecay$} 
\Output{Stationary point, $\genstateArgnt{n}$ }
\textbf{Online Steps:}\\ 
$\genstate^n_0 \leftarrow \bz$ \Comment{Set initial guess for state} \\
$\controllerArgnt{n} \leftarrow \boldsymbol 0$ \Comment{Set initial
guess for costate}\\ 
$\genstateArgnt{n}_1 \solveAlgo \basisspace^T \stweightingMatArg{n}
\basisspace \genstateDotArgnt{n}_1  -  \basisspace^T \stweightingMatArg{n}
\velocity(\veloargsromntArg{1}) =  \controllerArgnt{n}$ 
% Eq.~\ref{eq:lspg_continuous} with $\adjoint^n(t) = \controllerArg{n}{t}$ to
% obtain $\genstate_0^{n}(t)$ 
\Comment{Solve~\eqref{eq:lspg_continuous} with initial guess to
controller}\\ 
%Set: $\genstate^n_1 = \genstate$ \Comment{Set state} \\
%$\objectiveArg{n}({\genstate_1^n(t)})$
%\Comment{Evaluate objective function}\\
$i \leftarrow 1$ \Comment{Set counter}\\
\While{$tol \le \int_{\timeStartArg{n}}^{\timeEndArg{n}} \norm{\genstate^n_{i}(t) - \genstate^n_{i-1}(t) }dt $}{
\small{
\begin{multline*}
\adjointArgnt{n} \underset{\text{solve}}{\leftarrow} 
\frac{d}{dt} \adjoint^n  + \basisspace^T \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\basisspace \genstate^n_i + \stateIntercept,t) \bigg]^T \stweightingMatArg{n} \basisspace \mass^{-1} \adjoint^n = \\ -\bigg[\basisspace^T \bigg[ \frac{\partial \velocity}{\partial \stateyDiscrete} ( \basisspace \genstate^n_i + \stateIntercept,t) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} -   \basisspace \mass^{-1} \basisspace^T \bigg)  \stweightingMatArg{n} \bigg( \basisspace \dot{\genstate}_i^n   -   \velocity( \basisspace \genstate_i^n  +\stateIntercept,t) \bigg) \bigg] 
\end{multline*} }
\Comment{Solve~\eqref{eq:lspg_adjoint} to obtain guess to costate state} \\
%Solve Eq.~\ref{eq:lspg_adjoint} with $\genstate^n(t) = \genstate^n_i(t)$ to obtain $\adjoint^n(t)$ \\
$\controllerArgnt{n}  \leftarrow \rho \controllerArgnt{n} + (1 - \rho) \adjointArgnt{n}$ \Comment{Weighted update to costate}\\
$i \leftarrow i+1$ \Comment{Update Counter}\\
$\genstateArgnt{n}_i \solveAlgo \basisspace^T \stweightingMatArg{n} \basisspace \genstateDotArgnt{n}_i   -  \basisspace^T \stweightingMatArg{n} \velocity(\basisspace \genstate^n_i + \stateIntercept,t) =  \controllerArgnt{n} $
\Comment{Solve~\ref{eq:lspg_continuous} with updated guess to costate}\\
%$\objectiveArg{n}({\genstate_i^n(t)})$ \Comment{Evaluate objective function}\\
\uIf{ $\objectiveArg{n}({\genstate_i^n(t)}) \le \objectiveArg{n}({\genstate_{i-1}^n(t)})$}
{
$\rho \leftarrow \text{min}(\rho \fbsmGrowth,1)$ \Comment{Grow the damping factor}\\
}
\Else{
$\rho \leftarrow \frac{\rho }{ \fbsmDecay}$ \Comment{Shrink the damping factor}\\ 
$\genstate_i^{n} \leftarrow  \genstate_{i-1}^{n}$ \Comment{Reset state to value at previous iteration}
}
}
Return converged solution, $\genstateArgnt{n}= \genstateArgnt{n}_i$
\end{algorithm}
\subsubsection{Numerical considerations for the forward and backward problems}
The FBSM requires solving~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}, both of which are defined at the time-continuous level. 
The numerical implementation of the FBSM requires two main ingredients: (1.) temporal discretization schemes for the forward and backward problems and (2.) 
an efficient method for computing action of the Jacobian transpose on a vector.
